---
title: "Tidying and Wrangling"
format: html
---

```{r}
#| message: false

library(dplyr)
library(tidyr)
library(readr)

```

# Data sources

## Read in our data

```{r}
catch_original <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")

```

## Explore

```{r}
colnames(catch_original)

head(catch_original)

summary(catch_original)

unique(catch_original$Region)
```

Mac %>% shortcut: cmd + shift + M 

# Cleaning and Reshaping

## Using the select() function

```{r}

# select() to keep the columns that we want
catch_data <- catch_original %>% 
  select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum)
head(catch_data)

# select() to drop columns we don't want (-)
catch_data <- catch_original %>% 
  select(-All, -notesRegCode)
```

## Check column types

```{r}

glimpse(catch_data)

```

## Change column contents with mutate()

```{r}

# Change Chinook column to numeric
catch_clean <- catch_data %>% 
  mutate(Chinook = as.numeric(Chinook))

```


```{r}

# get the location of the NA value
i <- which(is.na(catch_clean$Chinook))
i

# look at the row in the data 
catch_data[i,]

# use mutate to fix
catch_clean <- catch_data %>% 
  mutate(Chinook = if_else(condition = Chinook == "I", 
                           true = "1",
                           false = Chinook),
         Chinook = as.numeric(Chinook))

head(catch_clean)

catch_clean[i,]
                           
```

## Changing file shape using pivot_loger() and pivot_wider()

```{r}

catch_long <- catch_clean %>% 
  pivot_longer(cols = -c(Region, Year), 
               names_to = "species",
               values_to = "catch")
head(catch_long)

catch_wide <- catch_long %>% 
  pivot_wider(names_from = species,
              values_from = catch)

head(catch_wide)

```


## Rename columns with rename()

```{r}

catch_long <- catch_long %>% 
  rename(catch_thousands = catch)

head(catch_long)

```

```{r}

catch_long <- catch_long %>% 
  mutate(catch = catch_thousands * 1000) %>% 
  select(-catch_thousands)

head(catch_long)

```

## Summary statistics using group_by and summarize()

```{r}

# Get average catch by region

mean_region <- catch_long %>% 
  group_by(Region) %>% 
  summarize(mean_catch = mean(catch))

head(mean_region)

# Get number of rows per group

n_region <- catch_clean %>% 
  group_by(Region) %>% 
  summarize(n = n())

head(n_region)

```

# Filtering rows using filter()

```{r}

sse_catch <- catch_long %>% 
  filter(Region == "SSE")

head(sse_catch)

catch_million <- catch_long %>% 
  filter(catch >= 1000000)

head(catch_million)

chinook_sse <- catch_long %>% 
  filter(Region == "SSE" & species == "Chinook")

```

## Sorting data with arrange()

```{r}

mean_region <- catch_long %>% 
  group_by(Region) %>% 
  summarize(mean_catch = mean(catch)) %>% 
  arrange(mean_catch)

head(mean_region)

# Decending order next 

mean_region <- catch_long %>% 
  group_by(Region) %>% 
  summarize(mean_catch = mean(catch)) %>% 
  arrange(desc(mean_catch))

head(mean_region)

```

## Combining datasets together with join()

```{r}

region_defs <- read.csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.303.1") %>% 
    select(code, mgmtArea)

head(region_defs)

catch_joined <- left_join(catch_long, region_defs, by = c("Region" = "code"))

head(catch_joined)

```

```{r}

region_defs <- read.csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.303.1") %>% 
    select(code, mgmtArea) %>% 
  rename(Region = code, Region_Name = mgmtArea)

catch_joined <- left_join(catch_long, region_defs, by = c("Region"))

head(catch_joined)

```

## Split and combine values in columns 

```{r}

sites_df <- data.frame(site = c("HAW-101", "HAW-103", "OAH-320"))
 
head(sites_df)

split_df <- sites_df %>% 
  separate(site, c("island", "site_number"), sep = "-")

head(split_df)

```

```{r}

dates_df <- data.frame(
  year = c("1930", "1930", "1932"),
  month = c("12", "12", "12"),
  day = c("14", "15", "16"))

head(dates_df)

dates_unite <- dates_df %>% 
  unite(col = date, year, month, day, sep = "-")

head(dates_unite)

```

```{r}

mean_region <- catch_original %>% 
  select(-All, -notesRegCode) %>% 
  mutate(Chinook = if_else(Chinook == "I",
                           "1",
                           Chinook)) %>% 
  mutate(Chinook = as.numeric(Chinook)) %>% 
  pivot_longer(-c(Region, Year),
               names_to = "species",
               values_to = "catch") %>% 
  mutate(catch = catch * 1000) %>% 
  group_by(Region) %>% 
  summarize(mean_catch = mean(catch)) %>% 
  arrange(desc(mean_catch))

head(mean_region)

```

## save results with write_csv()

```{r}

library(here)
write_csv(mean_region, here("data/mean_catch_by_region.csv"))

```


